A preliminary constraint grammar for Russian

### 2015-03-24 2015-nodalida-ruscg/ruscg.tex
added ambiguity intro
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@59500 72bbbca6-d526-0410-a7d9-f06f51895060

\section{Review of literature}
% ROB Russian POS tagging

@@

State-of-the-art morphological analysis in Russian is primarily based on finite-
state technology \cite{Nozhov-03,Segalovich-03}.\footnote{Machine-learning approaches have also been successfully applied to Russian, most notably by \newcite{sharoff08lrec-mocky}.} Almost without exception, all large-scale morphological transducers of Russian are based on the
forward-looking \emph{Grammatical Dictionary of Russian} \cite{Zaliznjak-77}.
This dictionary gives fine-grained morphological specifications for more than
100~000 words, including inflectional endings, morphophonemic alternations, 
stress patterns, exceptions, and idiosyncratic collocations.
We base our work on the recently developed [anonymous] morphological transducer, which is also based on Zaliznjak's dictionary (Author, 2015).\footnote{The [anonymous] transducer is implemented using a two-level 
morphology \cite{Koskenniemi-84}, and can be compiled using either {\tt xfst}
\cite{Beesley.Karttunen-03} or {\tt hfst} \cite{hfst-11}} This finite-state transducer (FST) generates all possible 
morphosyntactic readings of each wordform, regardless of its frequency or probability. Because Russian is a relatively highly inflected language, broad coverage is important, but widespread homonymy leads to the generation of many spurious readings, as discussed in Section~\ref{ambiguity} below. Because of this, one of the foundational steps in Russian natural language processing is morphosyntactic disambiguation, or \emph{snjatie omonimii}.

@@

% ROB
% homography

% sketch out the different kinds of ambiguity and show what percentage they are 
% give examples
% describe stress related ambiguity, e.g. where orthographic form is the same but diff. stress

\begin{table}
  \centering
  \begin{tabular}{|l|r|r|}
    \hline
    \textbf{Type}     & \textbf{\% all tokens} & \textbf{\% ambig. tokens} \\
    \hline
    Intraparadigmatic & 59.0                   & 90.9   \\
    Incongruent       & 27.7                   & 42.7   \\ 
    Congruent         & 1.20                   & 1.8    \\ 
    \hline
  \end{tabular}
  \caption{Types of ambiguity}
  \label{table:ambiguity}
\end{table}

>>>>

\section{Review of literature}

State-of-the-art morphological analysis in Russian is primarily based on
finite-state technology
\cite{Nozhov-03,Segalovich-03}.\footnote{Machine-learning approaches have also been successfully applied to Russian, most notably by 
\newcite{sharoff08lrec-mocky}.} Almost without exception, all large-scale 
morphological transducers of Russian are based on the
forward-looking \emph{Grammatical Dictionary of Russian} \cite{Zaliznjak-77}.
This dictionary gives fine-grained morphological specifications for more than
100~000 words, including inflectional endings, morphophonemic alternations, 
stress patterns, exceptions, and idiosyncratic collocations.
We base our work on the recently developed [anonymous] morphological transducer, which is also based on Zaliznjak's dictionary (Author, 2015).\footnote{The [anonymous] transducer is implemented using a two-level 
morphology \cite{Koskenniemi-84}, and can be compiled using either {\tt xfst}
\cite{Beesley.Karttunen-03} or {\tt hfst} \cite{hfst-11}} This finite-state transducer (FST) generates all possible 
morphosyntactic readings of each wordform, regardless of its frequency or probability. Because Russian is a relatively highly inflected language, broad coverage is important, but widespread homonymy leads to the generation of many spurious readings, as discussed in Section~\ref{ambiguity} below. Because of this, one of the foundational steps in Russian natural language processing is morphosyntactic disambiguation, or \emph{snjatie omonimii}.

@@

@@

We identify three different types of morphosyntactic ambiguity: 
intraparadigmatic, morphosyntactically incongruent, and morphosyntactically 
congruent. A recent study by Author (2015) found that about 7.5\% of 
morphosyntactic ambiguity in a corpus of Russian resulted in stress placement 
ambiguity. The following examples make use of stress ambiguity to illustrate 
each kind of ambiguity. \emph{Intraparadigmatic} ambiguity refers to homographic 
wordforms belonging to the same lexeme, as shown in \ref{ex:intrahom}.

\needspace{4\baselineskip} % keeps example all together (no orphan line) 
\ex. Intraparadigmatic homographs \label{ex:intrahom}
\a. \rus{т\'{е}ла} \emph{t\'{e}la} `body.\textsubscript{SG-GEN}' 
    \label{ex:bodySGGEN}
\b. \rus{тел\'{а}} \emph{tel\'{a}} `body.\textsubscript{PL-NOM}' 
    \label{ex:bodyPLNOM}

The remaining two types of ambiguity occur between lexemes. 
\emph{Morphosyntactically incongruent} ambiguity occurs between homographs that 
belong to separate lexemes, and whose morphosyntactic values are different, as 
shown in \ref{ex:MSincongruent}.

\needspace{4\baselineskip} % keeps example all together (no orphan line) 
\ex. Morphosyntactically incongruent homographs \label{ex:MSincongruent}
\a. \rus{н\'{а}шей} \emph{nášej} `our.\textsubscript{F-SG-GEN/DAT/LOC...}'\\
    \rus{наш\'{е}й} \emph{našéj} `sew on.\textsubscript{IMP-2SG}'
\b. \rus{дор\'{о}га} \emph{doróga} `road.\textsubscript{N-F-SG-NOM}'\\
    \rus{дорог\'{а}} \emph{dorogá} `dear.\textsubscript{ADJ-F-SG-PRED}'

\emph{Morphosyntactically congruent} ambiguity occurs between homographs 
that belong to separate lexemes, and whose morphosyntactic values are identical, 
as shown in \ref{ex:MScongruent}. As indicated in Table~\ref{table:ambiguity}, 
this kind of ambiguity is relatively rare, and resolving this ambiguity requires 
the use of technologies such as Word Sense Disambiguation.

\needspace{6\baselineskip} % keeps example all together (no orphan line) 
\ex. Morphosyntactically congruent homographs \label{ex:MScongruent}
\a. \rus{з\'{a}мок} \emph{z\'{a}mok} `castle.\textsubscript{SG-NOM}'\\
	\rus{зам\'{о}к} \emph{zam\'{o}k} `lock.\textsubscript{SG-NOM}'
\b. \rus{з\'{a}мков} \emph{z\'{a}mkov} `castle.\textsubscript{PL-GEN}'\\
	\rus{замк\'{о}в} \emph{zamk\'{o}v} `lock.\textsubscript{PL-GEN}'
\c. ...\\
	...

Table \ref{table:ambiguity} shows the proportional prevalence of each kind of 
ambiguity. Note that these proportions do not sum to 1, since a given token may 
exhibit more than one kind of ambiguity. The first column shows the proportion 
of all tokens in a corpus that have each kind of ambiguity. The second column 
shows what proportion of ambiguous tokens exhibit each kind of ambiguity.

Table~\ref{table:ambiguity} shows that most morphosyntactic ambiguity in 
unrestricted Russian text is rooted in intraparadigmatic and morphosyntactically 
incongruent ambiguity. Detailed part-of-speech tagging with morphosyntactic 
analysis can help disambiguate these forms. Morphosyntactically congruent stress 
ambiguity cannot be disambiguated by means of detailed part-of-speech tagging, 
and since congruent stress ambiguity represents only a small percentage of 
ambiguous wordforms, we leave it to future work.

\begin{table}
  \centering
  \begin{tabular}{l|rr}
    \hline
    \textbf{Type}  & \textbf{all tokens} & \textbf{ambig. tokens} \\
    \hline
    Intraparadigm. & 0.590                   & 0.909   \\
    Incongruent    & 0.277                   & 0.427   \\ 
    Congruent      & 0.120                   & 0.180    \\ 
    \hline
  \end{tabular}
  \caption{Proportional frequency of different types of morphosyntactic ambiguity in unrestricted text}
  \label{table:ambiguity}
\end{table}
